Statistical methods for Mendelian randomization in genome-wide association studies: A review

### 2022-04-28 reply-to-review1/mr-review.Rmd
fixed title

---
title: "Mendelian randomization with genome-wide association studies data"
author: "Frederick J. Boehm & Xiang Zhou"
date: "`r lubridate::now()`"
csl: ../elsevier-with-titles.csl
link-citations: yes
linkcolor: blue
always_allow_html: yes
bibliography: ../exported.bib
output:
  bookdown::word_document2:
    number_sections: false
---

@@

The GMM estimate is the value of the vector $\beta$ that satisfies Equation \@ref(eq:gmm), where $f(x_i; \beta) = \mathbb{E}(Y|\text{do}(X = x))$. @palmer_instrumental_2011 developed numerical methods for obtaining estimates from Equations \@ref(eq:gmm).

@@

It is worthwhile to consider ways in which the standard MR assumptions may be violated and strategies for identifying such violations. Let’s first consider the exclusion restriction assumption (assumption 2 above). It states that the SNP instrument must affect the outcome only through the exposure. The causal diagram in Figure 3B illustrates a scenario that violates this assumption [@vanderweele_methodological_2014]. Bias in causal effect estimation would result. However, if the intermediate is measured and treated as the exposure, then an unbiased estimate of causal effect is possible, since the intermediate captures the two pathways, intermediate -> outcome and intermediate -> exposure -> outcome, through which the SNP instrument affects the outcome Y [@vanderweele_methodological_2014].

```{r, dags, fig.cap = "Causal diagrams for MR. A. Scenario where the genetic variant affects an exposure, which in turn affects an outcome. B. Scenario where the genetic variant affects an intermediate variable on the pathway to the exposure. Because the intermediate affects the outcome through a pathway that doesn’t involve exposure, this scenario violates the exclusion restriction assumption. C. Correlated horizontal pleiotropy occurs when the genetic variant affects the exposure, which in turn affects the outcome, and the genetic variant affects the unobserved confounder, which in turn affects the exposure and the outcome independently."}
knitr::include_graphics("../figures/all_dags-two.png")
```

@@

Other MR methods can be used in sensitivity analysis to assess robustness of findings to 
possible violations of MR assumptions.
Additionally, methods like MR-clust have the potential to reveal additional relationships among genes that share a biological pathway [@foley_mr-clust_2021]. The STROBE-MR statement has additional practice recommendations [@skrivankova_strengthening_2021].

@@

Open research questions remain in the MR field. We have highlighted above three recent findings: omnigenic MR, sample overlap in biobanks, and bidirectional causal inference. While we discussed initial findings in these three areas, many opportunities for methods enhancements remain. For example, as biobank sizes grow in number of subjects and number of measured traits, there is increasing demand for MR methods that scale efficiently with large sample sizes. Development of methods for two-sample MR with overlapping samples opens opportunities to study rich biobank data, and we anticipate that this question - how to accommodate overlapping samples - will continue to be an active area of research.

>>>>

---
title: "Statistical methods for Mendelian randomization in genome-wide association studies: A review"
author: "Frederick J. Boehm & Xiang Zhou"
date: "`r lubridate::today()`"
csl: ../elsevier-with-titles.csl
link-citations: yes
linkcolor: blue
always_allow_html: yes
bibliography: ../exported.bib
output:
  bookdown::word_document2:
    number_sections: false
---

@@

The GMM estimate is the value of the vector $\beta$ that satisfies Equation \@ref(eq:gmm), where $f(x_i; \beta) = \mathbb{E}(Y|\text{do}(X = x))$. @palmer_instrumental_2011 developed numerical methods for obtaining estimates from Equation \@ref(eq:gmm).

@@

It is worthwhile to consider ways in which the standard MR assumptions may be violated and strategies for identifying such violations. Let’s first consider the exclusion restriction assumption (assumption 2 above). It states that the SNP instrument must affect the outcome only through the exposure (Figure 3B). The causal diagram in Figure 3A illustrates a scenario that violates this assumption [@vanderweele_methodological_2014]. Bias in causal effect estimation would result. However, if the intermediate is measured and treated as the exposure, then an unbiased estimate of causal effect is possible, since the intermediate captures the two pathways, intermediate -> outcome and intermediate -> exposure -> outcome, through which the SNP instrument affects the outcome Y [@vanderweele_methodological_2014].

```{r, dags, fig.cap = "Causal diagrams for MR. A. Scenario where the genetic variant affects an intermediate variable on the pathway to the exposure. Because the intermediate affects the outcome through a pathway that doesn’t involve exposure, this scenario violates the exclusion restriction assumption. B. Scenario where the genetic variant affects an exposure, which in turn affects an outcome, possibly in the presence of unmeasured confounding. C. Correlated horizontal pleiotropy occurs when the genetic variant affects the exposure, which in turn affects the outcome, and the genetic variant affects the unobserved confounder, which in turn affects the exposure and the outcome independently."}
knitr::include_graphics("../figures/all_dags.png")
```

@@

Other MR methods can be used in sensitivity analysis to assess robustness of findings to 
possible violations of MR assumptions.
Methods like MR-clust have the potential to reveal additional relationships among genes that share a biological pathway [@foley_mr-clust_2021]. The STROBE-MR statement has additional practice recommendations [@skrivankova_strengthening_2021].

@@

Open research questions remain in the MR field. We have highlighted above three recent findings: omnigenic MR, methods for sample overlap in biobanks, and methods for bidirectional causal inference. While we discussed initial findings in these three areas, many opportunities for methods enhancements remain. For example, as biobank sizes grow in number of subjects and number of measured traits, there is increasing demand for MR methods that scale efficiently with large sample sizes. Development of methods for two-sample MR with overlapping samples opens opportunities to study rich biobank data, and we anticipate that this question - how to accommodate overlapping samples - will continue to be an active area of research.
